---
title: "Data Wrangling"
author: "Rodrigo Girão Serrão"
date: "30 de Outubro de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Start by finding the current working directory and change it. Find the location of a file we will be using to test things.

```{r}
setwd("H:/edx-DataScience/c6-data-wrangling")
getwd()

path <- system.file("extdata", package="dslabs")
list.files(path)

if (! file.exists("murders.csv")) {
  full_path <- file.path(path, "murders.csv")
  file.copy(full_path, getwd())
  file.exists("murders.csv")
} else {
  print("no need to copy the murders.csv again")
}
```

Now we will load the data. (notice it has a header)

```{r}
library(tidyverse)
read_lines("murders.csv", n_max=3)
data <- read_csv("murders.csv")
head(data)

# try the base functions as well
data2 <- read.csv("murders.csv")
class(data2)
class(data2$abb)
data3 <- read.csv("murders.csv", stringsAsFactors = FALSE)
class(data3$abb)
```

Alternatively, we could've downloaded the data from the internet if we didn't have it locally:

```{r}
# define the url
url <- "https://raw.githubusercontent.com/rafalab/dslabs/master/inst/extdata/murders.csv"
# create a temporary file
tmp_filename <- tempfile()
download.file(url, tmp_filename)
data <- read_csv(tmp_filename)
file.remove(tmp_filename)
```

Check how one of the original files was and notice it was *not* tidy; use `gather` to make it tidy; convert the year to integers.

```{r}
path <- system.file("extdata", package="dslabs")
filename <- file.path(path, "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
head(wide_data)

new_tidy_data <- wide_data %>%
                gather(year, fertility, '1960':'2015')
new_tidy_data <- wide_data %>%
                gather(year, fertility, -country, convert=TRUE)

new_tidy_data %>% spread(year, fertility) # goes back
```

Off to wrangling something else!

```{r}
path <- system.file("extdata", package="dslabs")
filename <- file.path(path, "life-expectancy-and-fertility-two-countries-example.csv")

raw_data <- read_csv(filename)
select(raw_data, 1:5)

data <- raw_data %>% gather(key, value, -country)
head(data)

library(reader)

# the "extra=merge" is to handle the cases of year_life_expectancy, because those have 2 underscores
data %>% separate(key, c("year", "variable_name"), sep="_", extra="merge") %>%
    spread(variable_name, value)

# and now a not-so-decent way of doing the same thing
data %>% separate(key, c("year", "first_variable_name", "second_variable_name"), fill="right")
  
data %>% separate(key, c("year", "first_variable_name", "second_variable_name"), fill="right") %>% # fill the right variable with NAs (because of life expectancy)
    unite(variable_name, first_variable_name, second_variable_name, sep="_") %>% # unite those columns
    spread(variable_name, value) %>%
    rename(fertility = fertility_NA)
```

Now we see how we can join several tables. Take subsets of these two different tables and combine them in different ways.

```{r}
library(dslabs)
data(murders)
data(polls_us_election_2016)

tab1 <- slice(murders, 1:6) %>% select(state, population)
tab1
tab2 <- slice(results_us_election_2016, c(1:3,5,7:8)) %>%
        select(state, electoral_votes)
tab2

# add the electoral votes for the states we can
left_join(tab1, tab2)
tab1 %>% left_join(tab2)
right_join(tab1, tab2)
left_join(tab2, tab1)
inner_join(tab1, tab2)
full_join(tab1, tab2)

# keep the parts of the 1st table for which we have info on the second
semi_join(tab1, tab2)
# keep the parts of the 1st table for which we have NO info on the second
anti_join(tab1, tab2)
```

Let us scrape Wikipedia for a table.

```{r}
# this lib is in tidyverse as well
library(rvest)
url <- "https://en.wikipedia.org/wiki/Murder_in_the_United_States_by_state"
h <- read_html(url)
class(h)

tab <- h %>% html_nodes("table")
tab <- tab[[2]] %>% html_table
tab
```

### Some string processing

```{r}
v <- tab[,2]
str_detect(v, ",")
v <- str_replace_all(v, ",", "")
str_detect(v, ",")
v <- as.numeric(v)
class(v)
```

Use regex to correct some heights not in inches

```{r}
data("reported_heights")
rh <- reported_heights
pattern <- "^\\d\'[ ]*\\d{1,2}\"$"
sum(str_detect(rh$height, pattern))

# now try replacing some other representations to make
# everything more uniform: x'y
pattern <- "^\\d\'[ ]*\\d{1,2}$"
rh$height %>% str_replace("feet|ft|foot","'") %>%
        str_replace("inches|in|''|\"","") %>%
        str_detect(pattern) %>% sum
```

Now put everything together and clean the heights from scratch:

```{r}
pattern <- "^([4-7])\\s*'\\s*(\\d+\\.?\\d*)$"

convert_format <- function(s){
  s %>%
    str_replace("feet|foot|ft", "'") %>% #convert feet symbols to '
    str_replace_all("inches|in|''|\"|cm|and", "") %>%  #remove inches and other symbols
    str_replace("^([4-7])\\s*[,\\.\\s+]\\s*(\\d*)$", "\\1'\\2") %>% #change x.y, x,y x y
    str_replace("^([56])'?$", "\\1'0") %>% #add 0 when to 5 or 6
    str_replace("^([12])\\s*,\\s*(\\d*)$", "\\1\\.\\2") %>% #change european decimal
    str_trim() #remove extra space
}

words_to_numbers <- function(s){
  str_to_lower(s) %>%  
    str_replace_all("zero", "0") %>%
    str_replace_all("one", "1") %>%
    str_replace_all("two", "2") %>%
    str_replace_all("three", "3") %>%
    str_replace_all("four", "4") %>%
    str_replace_all("five", "5") %>%
    str_replace_all("six", "6") %>%
    str_replace_all("seven", "7") %>%
    str_replace_all("eight", "8") %>%
    str_replace_all("nine", "9") %>%
    str_replace_all("ten", "10") %>%
    str_replace_all("eleven", "11")
}

smallest <- 50
tallest <- 84
new_heights <- reported_heights %>% 
  mutate(original = height, 
         height = words_to_numbers(height) %>% convert_format()) %>%
  extract(height, c("feet", "inches"), regex = pattern, remove = FALSE) %>% 
  mutate_at(c("height", "feet", "inches"), as.numeric) %>%
  mutate(guess = 12*feet + inches) %>%
  mutate(height = case_when(
    !is.na(height) & between(height, smallest, tallest) ~ height, #inches 
    !is.na(height) & between(height/2.54, smallest, tallest) ~ height/2.54, #centimeters
    !is.na(height) & between(height*100/2.54, smallest, tallest) ~ height*100/2.54, #meters
    !is.na(guess) & inches < 12 & between(guess, smallest, tallest) ~ guess, #feet'inches
    TRUE ~ as.numeric(NA))) %>%
  select(-guess)

new_heights %>%
  filter(not_inches(original)) %>%
  select(original, height) %>% 
  arrange(height) %>%
  View()

new_heights %>% arrange(height) %>% head(n=7)
```