---
title: "Gapminder data Exploration"
author: "Rodrigo Girão Serrão"
date: "24 de Outubro de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Gapminder data

Load our data and plot life expectancies/birth rates from 50 years ago and from 2012

```{r load_data}
library(dslabs)
library(ggplot2)
library(dplyr)
data(gapminder)
head(gapminder)

gapminder %>% filter(year == 1962) %>%
          ggplot(aes(fertility, life_expectancy, color=continent)) + geom_point()

gapminder %>% filter(year %in% c(1962, 2012)) %>%
          ggplot(aes(fertility, life_expectancy, color=continent)) + geom_point() +
          facet_grid(continent~year)
```

Notice how the majority of the continents evolved greatly
```{r evolution}
gapminder %>% filter(year %in% c(1962, 2012)) %>%
          ggplot(aes(fertility, life_expectancy, color=year)) + geom_point() +
          facet_wrap(~continent)
```

```{r time_evolution}
gapminder %>% group_by(continent, year) %>% summarize(le = mean(life_expectancy)) %>%
              ggplot(aes(year, le, color=continent)) +
              geom_line()
```

Now we study the income (in dollars per day per person) accross different regions, for 1970
```{r income_per_day_per_person}
gapminder %>% filter(year == 1970 & !is.na(gdp)) %>%
              mutate(dollars_per_day = gdp/population/365) %>% 
              mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%
              ggplot(aes(region, dollars_per_day, fill = continent)) +
              geom_boxplot() +
              theme(axis.text.x = element_text(angle=90, hjust=1)) +
              scale_y_continuous(trans="log2")
```

Find groups for which data is available for 1970 and 2010 and divide the world into "west" and "developing", making two groups.
```{r}
list1 <- gapminder %>% filter(year == 1970 & !is.na(gdp)) %>% .$country
list2 <- gapminder %>% filter(year == 2010 & !is.na(gdp)) %>% .$country
countries <- intersect(list1, list2)
length(countries)
west_regions <- c("Southern Europe","Polynesia", "Western Europe",
                  "Australia and New Zealand", "Northern America", "Northern Europe")
gapminder %>% filter(country %in% countries & year %in% c(1970, 2010)) %>%
              mutate(group = ifelse(region %in% west_regions, "West", "Developing")) %>%
              mutate(dollars_per_day = gdp/population/365) %>% 
              ggplot(aes(dollars_per_day)) +
              geom_histogram(binwidth=1, color="black") + 
              scale_x_continuous(trans="log2") +
              facet_grid(year~group)
```

A dictator ruled Cambodia between 1975 and 1979, look at his damage:
```{r dictator}
gapminder %>% filter(country=="Cambodia" & year >= 1960 & year <= 2010) %>%
        ggplot(aes(year, life_expectancy)) + geom_line()
```

and now a nice plot from the exercises in Data Camp.
```{r}
gapminder %>% filter(continent=="Africa" & year %in% c(1970, 2010) & !is.na(gdp)) %>% mutate(dollars_per_day = gdp/population/365) %>%
  ggplot(aes(dollars_per_day, fill=region)) + geom_density(bw=0.5, position="stack") + facet_grid(~year) + scale_x_continuous(trans="log2")
```