---
title: "Probability"
author: "Rodrigo Girão Serrão"
date: "26 de Outubro de 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Simple simulation

```{r}
B <- 10000
urn <- rep(c("blue", "red"), times=c(2,3))
tab <- table(replicate(B, sample(urn, 1)))/B

tab2 <- table(sample(urn, B, replace=TRUE))/B
```

Create a card deck
```{r}
suits <- c("diamonds", "hearts", "spades", "clubs")
values <- c("ace", "two", "three", "four", "five", "six", "seven", "eight", "nine", "ten", "jack", "queen", "king")
deck <- expand.grid(suit=suits, value=values)
deck <- paste(deck$value, deck$suit)
```

and compute the probability of taking a king knowing you just took a king from the deck as well as the probability of getting a blackjack:

```{r}
library(gtools)
hands <- permutations(52, 2, v=deck)
kings <- paste("king", suits)
first_card <- hands[,1]
second_card <- hands[,2]
sum(first_card %in% kings & second_card %in% kings) / sum(first_card %in% kings)

aces <- paste("ace", suits)
facecard <- c("king", "jack", "queen", "ten")
facecard <- expand.grid(value=facecard, suit=suits)
facecard <- paste(facecard$value, facecard$suit)
hands <- combinations(52, 2, v=deck)
mean(hands[,1] %in% aces & hands[,2] %in% facecard)
# monte carlo sim.
B <- 10000
results <- replicate(B, {
  hand <- sample(deck, 2)
  ((hand[1] %in% aces & hand[2] %in% facecard) | (hand[1] %in% facecard & hand[2] %in% aces))
})
mean(results)
```

Use Monte Carlo simulations to calculate the probability that, if n people are in a classroom, at least two of them share the same birthday.

```{r}
compute_prob <- function(n, B=10000) {
  mean(replicate(B, {
  bdays = sample(1:365, n, replace=TRUE)
  any(duplicated(bdays))
  }))
}

exact_prob <- function(n) {
  factors <- seq(365, 365-n+1)/365
  1 - prod(factors)
}

n <- 1:60
lookup_table <- sapply(n, compute_prob)
exact_table <- sapply(n, exact_prob)
lookup_table[40]
exact_table[40]
plot(n, lookup_table)
plot(n, exact_table)
```

A Monte Carlo simulation with normally distributed values. In particular, mimic the sample from the *heights* data set:

```{r}
library(dplyr)
library(dslabs)
library(ggplot2)
data(heights)
x <- heights %>% filter(sex=="Male") %>% .$height
n <- length(x)
avg <- mean(x)
s <- sd(x)
simulated_heights <- rnorm(n, avg, s)

all_heights <- c(x, simulated_heights)
group <- c(rep("real", n), rep("simulated", n))

data.frame(heights=all_heights, type=group) %>%
  ggplot(aes(heights, fill=type)) +
  geom_density(alpha=0.8, bw=1, position="identity")
```